---
title: 'Wharf to Wharf: 2018'
author: "Aaron Ferrucci"
date: "July 27, 2018"
output: html_document
---

### Introduction
The Wharf to Wharf race happens every July in Santa Cruz. The mostly flat,
6-mile course hugs the coastline between Santa Cruz and Capitola. The 
race has grown in popularity since its beginning in 1973, with more than 13,000
registered participants in 2018.

Race results are available online. You
can look up runners' results by bib number, name, city and state. This
is ok, but it's not very handy for anyone wanting to do real data analysis.

I've written some code to scrape the data from the results page, clean it
up, and plot some interesting things. The code is available on github; if you
see a problem, fork my repo, fix it, and do a pull request!

``` {r init, echo=F,message=FALSE}
library(dplyr)
library(ggplot2)
library(gridExtra)
source("w2w_utils.R")
```
``` {r read_data, echo=F}
allData <- getData(2018)
cleanData <- clean(2018, allData)
cleanDataWithAge <- cleanData[cleanData$age > 5,]
```

### Summary of results
* There were `r nrow(allData)` participants in the 2015 event
* The average age of a participant is about `r round(mean(cleanDataWithAge$age), digits=0)`.
* There are more female participants than male, with a F:M ratio of about `r round(nrow(allData[allData$sex == "F",]) / nrow(allData[allData$sex == "M",]), 1)`:1.
* Overwhelmingly, participants are from the USA. And almost all (`r round(100 * nrow(allData[allData$state == "CA",]) / nrow(allData), 1 )`%) are from California.
* There are some anomalies in the data - `r nrow(allData[allData$age == 0,])` participants with age 0, `r nrow(allData[allData$start < 8 * 3600 * 1000,])` participants who started before the official start time of 8:30AM (but maybe this is real data), etc. It's not the messiest data set I've seen.
* Males tend to be faster than females. This is especially pronounced in the top 100 ranks.

### Basic Info
Histogram of runner ages:

``` {r echo=F, message=F}
qplot(cleanDataWithAge$age, ylab = "", xlab = "age")
```

Histogram of genders:

``` {r echo=F, message=F}
qplot(cleanData$sex, ylab = "", xlab = "sex")
```

Country of origin. In 2018, we have the USA, Australia and Kenya:

``` {r echo=F, message=F}
table(cleanData$country)
```

#### Charts
``` {r echo=F, message=F}
elapsed_ticks <- seq(0, max(cleanDataWithAge$elapsed), 900000)
ggplot(cleanDataWithAge, aes(x = age, y = elapsed, color=sex)) +
  scale_x_continuous(breaks = seq(0, 100, 10)) + 
  scale_y_continuous(breaks = elapsed_ticks, labels = timestr(elapsed_ticks), name = "elapsed time (hh:mm:ss)") + 
  geom_point() +
  expand_limits(y = 0.375 * 3600 * 1000) +
  stat_smooth(method = "gam", formula = y ~ s(x, bs="cs"))
```

``` {r echo=F, message=F}
start_ticks <- seq(8.5 * 3600 * 1000, max(cleanData$start), 0.0625 * 3600 * 1000)
ggplot(cleanData, aes(x = elapsed, y = start, color = sex)) + 
  scale_y_continuous(breaks = start_ticks, labels = timestr(start_ticks)) + 
  scale_x_continuous(breaks = elapsed_ticks, labels = timestr(elapsed_ticks)) +
  expand_limits(x = 0.25 * 3600 * 1000, y = 8.5 * 3600 * 1000) +
  geom_point()
```

``` {r echo=F, message=F}
top100 <- cleanData[cleanData$oversex <= 100 & cleanData$oversex > 0,]
top100_ticks <- seq(0, max(top100$elapsed), (5/60) * 3600 * 1000)
ggplot(top100, aes(x = age, y = elapsed, color=sex)) +
  scale_x_continuous(breaks = seq(0, 100, 10)) + 
  scale_y_continuous(breaks = top100_ticks, labels = timestr(top100_ticks), name = "elapsed time (hh:mm:ss)") + 
  geom_point() +
  expand_limits(y = (27/60) * 3600 * 1000) +
  stat_smooth(method = "gam", formula = y ~ s(x, bs="cs"))
```


#### References
Github link: https://github.com/aaronferrucci/wharf2wharf  
Wharf to Wharf main page: http://www.wharftowharf.com/  
Wharf to Wharf results page: http://www.wharftowharf.com/results  
Definition of jogging: https://en.wikipedia.org/wiki/Jogging
